\chapter*{Acknowlegments}

%% Insira aqui seus Agradecimentos. Não esqueça de mencionar as Agências
%% Financiadoras.

\vfill

I want to start by thaking CAPES for funding my research and making this work possible. \\

My deepest gratitude goes to my advisor, Marcos Lopes, for his unwavering support and belief in me throughout the rollercoaster of my master's experience. Marcos, your steadfast patience, guidance, and encouragement have been my guiding lights. \\

I extend my heartfelt thanks to my former advisor, Cida Araújo, for introducing me to Cognitive Linguistics and Talmy's work on Spatial Semantics. This research would not have been possible without our collaboration on two invaluable PIBIC/CNPq projects during my undergraduate studies, which laid the groundwork for the work presented here. Cida, your teachings have been instrumental in shaping me into the researcher I am today. \\

To my comrads from the graduate program, you all know who you are. We have shared the highs and lows, frustrations and laughs, deadlines and memes, making this academic pursuit in Computational Linguistics a memorable one. \\

To my dear friends in São Paulo and Minas Gerais, you also know who you are. Your encouragement, support, and virtual presence have been a constant source of strength. \\

Last but not least, I would like to thank my family -- especially \textbf{my mother}. As the first in my family to study at a public university and pursue a graduate degree, achieving this milestone represents breaking a cycle of social injustice and fulfilling their unachieved dreams. Mom, thank you for always encouraging me to study and for helping me change my life through education.

\vfill